Extracting Information from AGN Variability

### 2017-03-31 Manuscript.tex
Re-work the Abstract and Section 3
Abstract text suggesting the physical interpretation of C-ARMA
processes has been moved into Section 3 and is now presented as
a possibility rather than our claim.

% Abstract of the paper
\begin{abstract}
AGN exhibit rapid, high amplitude stochastic flux variability across the entire electromagnetic spectrum on timescales ranging from hours to years. The cause of this variability is poorly understood. We present a Green's Function-based method for using variability to (1) measure the time-scales on which flux perturbations evolve and (2) characterize the driving flux perturbations. We model the observed light curve of an AGN as a linear differential equation driven by stochastic impulses. Physically, the impulses could potentially be attributable to local `hot-spots' in the accretion disk---in which case, the linear differential equation describes how the hot spots evolve and dissipate. In this picture, the Green's function of the linear differential equation is the impulse-response function of the accretion disk material to small perturbations of the luminous flux. If accretion disk hot-spots are the root cause of AGN variability, the powerspectrum of the driving stochastic impulses characterizes the timescales on which the hot-spots evolve. We analyze the light curve of the \Kepler AGN Zw 229-15 and find that the observed variability behavior can be modeled as a damped harmonic oscillator perturbed by a colored noise process. The model powerspectrum turns over on time-scale $385$~d. On shorter time-scales, the log-powerspectrum slope varies between $2$ and $4$, explaining the behavior noted by previous studies. We recover and identify both the $5.6$~d and $67$~d timescales reported by previous work. In the local hot-spot picture, these timescales represent the time-scale on which flux perturbations grow, and the time-scale on which flux perturbations decay back to the steady-state flux level respectively. We make the software package \href{https://github.com/AstroVPK/kali}{\textsc{k\={a}l\={i}}} used to study light curves using our method available to the community.
\end{abstract}

@@

Recently, \citet{Kelly14} suggested that stochastic variability in a large class of astrophysical sources may be modelled using Continuous-time AutoRegressive Moving Average (C-ARMA) processes. The attractiveness of using C-ARMA processes to model stochastic phenomenon stems from two factors: (1) C-ARMA processes directly specify property 2, i.e., the correlation structure of the light curve by providing a differential equation that governs how the light curve evolves over time. This alternative to providing a second-order statistic is powerful because it makes the light curve evolution more amenable to physical interpretation as is shown in this work. It also provides some idea of how to replace Gaussian distributions with more generalized $\alpha$-stable distributions that may be truer to reality \citep{BrockwellMarquardt05, BrockwellLinder09}; and (2) flexibility of the shape of their PSD. Since the PSD of a C-ARMA process is a rational function, it can be made arbitrarily complex to model a very wide range of phenomena. This flexibility is achieved by allowing C-ARMA processes to have multiple Autoregressive (quantified by $p$) and Moving Average (quantified by $q$) terms under the constraint $q < p$. If $\chi$ is a small flux perturbation from the mean flux level, then a C-ARMA($p$,$q$) process that models the time evolution of $\chi$ has the general form
\begin{multline}\label{eq:CARMAIntro}
\mathrm{d}^{p}\chi + \alpha_{1} \mathrm{d}^{p-1}\chi + \ldots + \alpha_{p-1} \mathrm{d}\chi + \alpha_{p} \chi = \\ \beta_{0} (\mathrm{d}W) + \beta_{1} \mathrm{d}(\mathrm{d}W) + \ldots + \beta_{p-1} \mathrm{d}^{p-1}(\mathrm{d}W)
\end{multline}
where $W$ is a \textit{Wiener} process and is responsible for introducing stochasticity into the behavior of $\chi$. The $\alpha$ and $\beta$ are constants that govern the behavior of the resulting process and have dimensions of various powers of time. In this paper, we present a Green's function-based method for understanding what a C-ARMA process tells us about accretion physics.

@@

Let the instantaneous flux emitted by an AGN accretion disk be $F(t)$. We may write $F(t) = F_{0} + \chi(t)$ where the $\chi)t)$ are perturbations to the mean flux, $F_{0}$. While the physical origins of AGN variability remain unclear, the phenomenon is known to be non-linear in the X-ray \citep{UMV05} suggesting that some non-linear equation(s) governs the behavior of $F(t)$. It is not necessarily the case that this equation is physically meaningful i.e. it may be purely phenomenological, arising out of the complex interactions of numerous physical phenomenon. Regardless, if the $\chi(t)$ are small (e.g. $\sim 10 \%$ in the optical), we can \textit{linearize} the non-linear equation that governs $F(t)$ and obtain a \textit{linear} equation for the perturbations $\chi(t)$. In appendix~\ref{sec:LHS}, we show how the linearization of an arbitrary non-linear equation for the flux $F(t)$ yields the left hand side (LHS) of the C-ARMA equation,
\begin{equation}\label{eq:LHS}
\mathrm{d}^{p}\chi + \alpha_{1} \mathrm{d}^{p-1}\chi + \ldots + \alpha_{p-1} \mathrm{d}\chi + \alpha_{p} \chi = u.
\end{equation}
Here, $u$ are input impulses that drive the behavior of the flux perturbations, $\chi$.

@@

The cause of perturbations, i.e. variability, in the flux emanating from an AGN is unknown, but we may construct a very general mathematical model for the form of the flux perturbations based on physical and statistical arguments. The region of space surrounding the central engine is vast. The inner edge of the accretion disk is thought to begin a few $r_{G} \sim 20$~AU ($M_{BH} = 10^{9} M_{\sun}$) away from the event horizon, while the outer edge is thought to lie at a distance of a few parsec from the event horizon. Contributions to the deviations in the flux most likely simultaneously arise from many individual locations in this system. Under the assumption that the distribution of energy released by each individual event is drawn from a distribution with finite second-moment, the central-limit-theorem assures us that the total energy introduced into the system at any instant will be drawn from a normal distribution, i.e., some form of Weiner noise. Due to the non-linearity of the physics involved, it is quite likely that correlations exist between the energy released into the disk at different moments in time. Therefore, we shall model the correlations by treating the driving process $u$ in equation~\eqref{eq:LHS} as a correlated Gaussian noise process. In appendix~\ref{sec:RHS}, we argue why the natural mathematical form of the RHS should be given by
\begin{equation}\label{eq:RHS}
u = \beta_{0} (\mathrm{d}W) + \beta_{1} \mathrm{d}(\mathrm{d}W) + \ldots + \beta_{q-1} \mathrm{d}^{q-1}(\mathrm{d}W) + \beta_{q} \mathrm{d}^{q}(\mathrm{d}W),
\end{equation}
with $q < p$. This form for the input stochastic impulses can be characterized by the power spectral density (PSD) of the impulses given by
\begin{equation}\label{eq:ImpulsePSD}
S_{uu}(\nu) = \frac{1}{2 \upi } |\beta(2 \upi \imath \nu)|^{2},
\end{equation}
where $\beta(z)$ is the moving average polynomial of equation~\eqref{eq:MACharPoly}. The PSD of the stochastic impulses tells about the timescales on which the stochastic impulses occur. Combining equation~\eqref{eq:LHS} with equation~\eqref{eq:RHS} yields the original C-ARMA process equation~\eqref{eq:CARMAIntro}. A C-ARMA process is a linear differential equation perturbed/driven by a random process. This driving process may be viewed as disturbances generated by MRI instabilities in the accretion disk. \textit{Thus we see that we can describe AGN luminosity variations in terms of a linear system driven by colored noise}.

Fitting the observed light curve of an AGN to the CARMA process of equation~\eqref{eq:CARMAIntro} tells us how flux perturbations evolve via the Green's function equation~\eqref{eq:GFSol}. Insight into the flux perturbation producing mechanism may be gained from the impulse-PSD equation~\eqref{eq:ImpulsePSD} by comparing it to predictions from simulations of turbulence \citep{EddySim15} or from thermodynamic considerations of the accretion disk material \citep{Mellen52}. Alternatively, the values of characteristic time-scales implied by the forms of the Green's function equation~\eqref{eq:GFSol} and impulse-PSD equation~\eqref{eq:ImpulsePSD} may be matched to theoretical time-scales pertinent to the generation and evolution of accretion disk instabilities by the MRI process. In the next section we discuss a simple C-ARMA process---the C-ARMA($2$,$1$) process---in detail and derive the characteristic time-scales present in this process as an example.

@@

We shall discuss the C-ARMA($2$,$1$) process in detail because it has strong physical motivations and is relatively simple to interpret. Notice that in the process of linearising the original non-linear differential equation of equation~\eqref{eq:FluxDE}, we have not changed the order of the highest derivative of the flux that appears in that equation; if the original equation governing the flux emitted by the accretion disk is of $p^{\mathrm{th}}$-order, the corresponding linearised C-ARMA process is also of $p^{\mathrm{th}}$-order. Most physical phenomena are governed by $1^{\mathrm{st}}$- and  $2^{\mathrm{nd}}$-order equations. Hence, we may expect to find that in the linear regime, AGN variability is well described a C-ARMA($1$,$0$) process or a C-ARMA($2$,$q$) process with $q < 2$.

>>>>

% Abstract of the paper
\begin{abstract}
AGN exhibit rapid, high amplitude stochastic flux variations across the entire electromagnetic spectrum on timescales ranging from hours to years. The cause of this variability is poorly understood. We present a Green's Function-based method for using variability to (1) measure the time-scales on which flux perturbations evolve and (2) characterize the driving flux perturbations. We model the observed light curve of an AGN as a linear differential equation driven by stochastic impulses. We analyze the light curve of the \Kepler AGN Zw 229-15 and find that the observed variability behavior can be modeled as a damped harmonic oscillator perturbed by a colored noise process. The model powerspectrum turns over on time-scale $385$~d. On shorter time-scales, the log-powerspectrum slope varies between $2$ and $4$, explaining the behavior noted by previous studies. We recover and identify both the $5.6$~d and $67$~d timescales reported by previous work using the Green's Function of the C-ARMA equation rather than by directly fitting the powerspectrum of the light curve. These are the timescales the time-scale on which flux perturbations grow, and on which flux perturbations decay back to the steady-state flux level respectively. We make the software package \href{https://github.com/AstroVPK/kali}{\textsc{k\={a}l\={i}}} used to study light curves using our method available to the community.
\end{abstract}

@@

Recently, \citet{Kelly14} suggested that stochastic variability in a large class of astrophysical sources may be modelled using Continuous-time AutoRegressive Moving Average (C-ARMA) processes. The attractiveness of using C-ARMA processes to model stochastic phenomenon stems from two factors: (1) C-ARMA processes directly specify property 2, i.e., the correlation structure of the light curve by providing a differential equation that governs how the light curve evolves over time. This alternative to providing a second-order statistic is powerful because it makes the light curve evolution more amenable to physical interpretation as is shown in this work. It also provides some idea of how to replace Gaussian distributions with more generalized $\alpha$-stable distributions that may be truer to reality \citep{BrockwellMarquardt05, BrockwellLinder09}; and (2) flexibility of the shape of their PSD. Since the PSD of a C-ARMA process is a rational function, it can be made arbitrarily complex to model a very wide range of phenomena. This flexibility is achieved by allowing C-ARMA processes to have multiple Autoregressive (quantified by $p$) and Moving Average (quantified by $q$) terms under the constraint $q < p$. If $\chi$ is a small flux perturbation from the mean flux level, then a C-ARMA($p$,$q$) process that models the time evolution of $\chi$ has the general form
\begin{multline}\label{eq:CARMAIntro}
\mathrm{d}^{p}\chi + \alpha_{1} \mathrm{d}^{p-1}\chi + \ldots + \alpha_{p-1} \mathrm{d}\chi + \alpha_{p} \chi = \\ \beta_{0} (\mathrm{d}W) + \beta_{1} \mathrm{d}(\mathrm{d}W) + \ldots + \beta_{p-1} \mathrm{d}^{p-1}(\mathrm{d}W)
\end{multline}
where $W$ is a \textit{Wiener process} and is responsible for introducing stochasticity into the behavior of $\chi$. The $\alpha$ and $\beta$ are constants that govern the behavior of the resulting process and have dimensions of various powers of time. In this paper, we present a Green's function-based method for understanding what a C-ARMA process tells us about accretion physics.

@@

Let the instantaneous flux emitted by an AGN accretion disk be $F(t)$. We may write $F(t) = F_{0} + \chi(t)$ where the $\chi(t)$ are perturbations to the mean flux, $F_{0}$. While the physical origins of AGN variability remain unclear, the phenomenon is known to be non-linear in the X-ray \citep{UMV05}, suggesting that a non-linear equation(s) governs the behavior of $F(t)$. It is not necessarily the case that this equation is physically meaningful i.e. it may be purely phenomenological, arising out of the complex interactions of numerous physical phenomenon. Regardless, if the $\chi(t)$ are small (e.g. $\sim 10 \%$ in the optical), we can \textit{linearise} the non-linear equation that governs $F(t)$ and obtain a \textit{linear} equation for the perturbations $\chi(t)$. In appendix~\ref{sec:LHS}, we show how the linearisation of an arbitrary non-linear equation for the flux $F(t)$ yields the left hand side (LHS) of the C-ARMA equation,
\begin{equation}\label{eq:LHS}
\mathrm{d}^{p}\chi + \alpha_{1} \mathrm{d}^{p-1}\chi + \ldots + \alpha_{p-1} \mathrm{d}\chi + \alpha_{p} \chi = u.
\end{equation}
Here, $u$ are input impulses that drive the behavior of the flux perturbations, $\chi$.

@@

The cause of perturbations, i.e. variability, in the flux emanating from an AGN is unknown, but we may construct a very general mathematical model for the form of the flux perturbations based on physical and statistical arguments. The region of space surrounding the central engine is vast. The inner edge of the accretion disk is thought to begin a few $r_{G} \sim 20$~AU ($M_{BH} = 10^{9} M_{\sun}$) away from the event horizon, while the outer edge is thought to lie at a distance of a few parsec from the event horizon. Contributions to the deviations in the flux most likely simultaneously arise from many individual locations in this system. Under the assumption that the distribution of energy released by each individual event is drawn from a distribution with finite second-moment, the central-limit-theorem assures us that the total energy introduced into the system at any instant will be drawn from a normal distribution, i.e., some form of Weiner noise. Due to the non-linearity of the physics involved, it is quite likely that correlations exist between the energy released into the disk at different moments in time. Therefore, we shall model the correlations by treating the driving process $u$ in equation~\eqref{eq:LHS} as a correlated Gaussian noise process. In appendix~\ref{sec:RHS}, we argue why the natural mathematical form of the RHS should be given by
\begin{equation}\label{eq:RHS}
u = \beta_{0} (\mathrm{d}W) + \beta_{1} \mathrm{d}(\mathrm{d}W) + \ldots + \beta_{q-1} \mathrm{d}^{q-1}(\mathrm{d}W) + \beta_{q} \mathrm{d}^{q}(\mathrm{d}W),
\end{equation}
with $q < p$. This form for the input stochastic impulses can be characterized by the power spectral density (PSD) of the impulses given by
\begin{equation}\label{eq:ImpulsePSD}
S_{uu}(\nu) = \frac{1}{2 \upi } |\beta(2 \upi \imath \nu)|^{2},
\end{equation}
where $\beta(z)$ is the moving average polynomial of equation~\eqref{eq:MACharPoly}. The PSD of the stochastic impulses tells about the timescales on which the stochastic impulses occur. Combining equation~\eqref{eq:LHS} with equation~\eqref{eq:RHS} yields the original C-ARMA process equation~\eqref{eq:CARMAIntro}. A C-ARMA process is a linear differential equation perturbed/driven by a random process. \textit{We see that C-ARMA processes attempt to describe AGN luminosity variations in terms of a linear system driven by colored noise}.

Physically, the impulses could potentially be attributable to local `hot-spots' in the accretion disk---in which case, the linear differential equation describes how the hot spots evolve and dissipate. In this picture, the Green's function of the linear differential equation is the impulse-response function of the accretion disk material to small perturbations of the luminous flux. The driving process may be viewed as disturbances generated by MRI instabilities in the accretion disk. If accretion disk hot-spots are the root cause of AGN variability, the powerspectrum of the driving stochastic impulses characterizes the timescales on which the hot-spots evolve.

%Fitting the observed light curve of an AGN to the CARMA process of equation~\eqref{eq:CARMAIntro} tells us how flux perturbations evolve via the Green's function equation~\eqref{eq:GFSol}. Insight into the flux perturbation producing mechanism may be gained from the impulse-PSD equation~\eqref{eq:ImpulsePSD} by comparing it to predictions from simulations of turbulence \citep{EddySim15} or from thermodynamic considerations of the accretion disk material \citep{Mellen52}. Alternatively, the values of characteristic time-scales implied by the forms of the Green's function equation~\eqref{eq:GFSol} and impulse-PSD equation~\eqref{eq:ImpulsePSD} may be matched to theoretical time-scales pertinent to the generation and evolution of accretion disk instabilities by the MRI process.

In the next section we discuss a simple C-ARMA process---the C-ARMA($2$,$1$) process---in detail and derive the characteristic time-scales present in this process as an example.

@@

We shall discuss the C-ARMA($2$,$1$) process in detail because it is relatively simple to work with and interpret. Second order differential equations also have strong physical motivations; $2^{\mathrm{nd}}$-order differential equations such as the wave equation govern many physical phenomena.

The process of linearising the original non-linear differential equation of equation~\eqref{eq:FluxDE} preserves the order of the original non-linear differential equation, i.e. no changes occur in the order of the highest derivative of the flux that appears in that equation. This means that if the original equation governing the flux emitted by the accretion disk is of $p^{\mathrm{th}}$-order, the corresponding linearised C-ARMA process is also of $p^{\mathrm{th}}$-order. %Hence, we may expect to find that in the linear regime, AGN variability is well described a C-ARMA($1$,$0$) process or a C-ARMA($2$,$q$) process with $q < 2$.
